Principles for Open Data Curation: A Case Study with the New York City 311 Service Request Data

### 2024-11-11 manuscript/manuscript.tex
Completed editing
Minor grammatical and sentence structure correction. Ready for additional review.

Figure~\ref{fig:SRcountbyAgency} provides a breakdown of SRs by
the responsible Agency, showing the cumulative 
percentage of SRs handled by each
Agency over a the 2\mbox{-}year time-frame. Note that the
distribution of SRs is heavily concentrated among a few key 
agencies. The six largest agencies are:

@@

\begin{figure}[tbp]
	\centering
	\includegraphics[width = \textwidth]{SRs_by_Agency.pdf}
  	\caption{SR counts by Agency with Cumulative Percentage}
	\label{fig:SRcountbyAgency}
\end{figure}

@@

We identify the presence of issues in the NYC 311 SR open dataset;
not attempting to solve them, but rather to highlight their scope, 
magnitude, and potential impact. We would recommend 
that a corrective solution should be undertaken only 
after further investigation as to the why and how such 
issues might come about, including  detailed discussions 
with the originating Agency as to whether or not it is an 
actual error, correct,  or holds some other status.

@@

\paragraph{Zip Codes:}
\label{sec:zipcodesissues}
All \texttt{incident\_zip} zip codes should 
validate against the USPS database, which contains 
44,173 valid zip codes. And while the \texttt{incident\_zip} field has 
only 0.07\% invalid entries, this still results in 4,374 errors; 
not insignificant. If you group the invalid entries by 
percentage by Agency, it mirrors the \textit{overall} 
breakdown of SRs by Agency, potentially 
indicating a systemic problem.

@@

\paragraph{Closed before Created: Negative Duration}
Citizens, NYC Government Officials, and Agencies use the \texttt{created\_date}(s) and 
\texttt{closed\_date}(s) to measure the \textit{duration} of SRs, 
a useful surrogate to measure an Agency's responsiveness. 
While duration is not directly present in the dataset, 
it can be easily computed as the difference between
\texttt{(closed\_date} and \texttt{created\_date}.  Duration is one of 
most frequently used 311metrics and is frequently analyzed, such as determining if 
one NYC borough receives faster response than another for certain 
complaints. There are 12,450 SRs where the \texttt{closed\_date} precedes the 
\texttt{created\_date}, generating nonsensical ``negative durations.'' 
Although this represents only 0.2\% of the dataset, these errors can 
significantly impact response time analyses. Eight SRs with extremely 
large negative durations ($-$730 days), all originating from the 
Department of Homeless Services (DHS); all contain an entry 
of ``1900-01-01'' as the \texttt{closed\_date}. This results 
in negative durations exceeding $-$44,601 days (or 122 years). Many
other SRs have negative durations in excess of 300+ days. Such 
anomalies, though rare, can significantly skew statistical results if 
not addressed during data cleansing. As a result, these SR rows are 
removed from the analysis.

@@

\paragraph{Identical Created Date \& Closed Date: Zero Durations}
A more prevalent issue occurs when the \texttt{closed\_date} and 
\texttt{created\_date} are exactly the same, to the second. This 
creates a \textbf{zero duration} for the SR, which is nonsensical. There 
are 163,720 zero duration SRs, representing 2.6\% of all non\mbox{-}blank data. 
99\% percent of these zero duration SRs occur within five agencies:
DHMH, DOT, DOB, DSNY, and DEP. This distribution \textbf{does not} 
mirror the overall SR distribution by Agency, suggesting 
an Agency\mbox{-}specific issue.

@@

\paragraph{\texttt{borough} \& \texttt{taxi\_company\_borough}:} Despite 
their names, these fields are almost entirely different, with only 
a 0.05\% match. The \texttt{taxi\_company\_borough} field is 
used exclusively by the TLC, indicating a need for consultation 
with that Agency to understand the differing uses. It is 
likely that that the usage of \texttt{taxi\_company\_borough} is 
quite different than that of \texttt{borough}.

\paragraph{\texttt{agency} \& \texttt{agency\_name}:} The \texttt{agency} 
field contains the abbreviations of City agencies (e.g., NYPD, DOT, DSNY), 
while \texttt{agency\_name} contains the full names. Including both seems 
redundant, especially given the space taken up by full names in the 
CSV file. Additionally, the Agency abbreviations are easily recognized.

@@

\subsection{Principles}
\paragraph{Principle 1: Ensuring Consistency Across Agencies and Time}
A key challenge in government released datasets is the 
harmonization of data fields across various agencies and 
maintaining this consistency over time. To address this, agencies should adopt 
standardized naming conventions and formats across all datasets. 
Regular audits and data harmonization processes should be 
implemented to ensure consistency across long\mbox{-}term datasets, 
particularly as agency structures or reporting standards evolve. 
For instance, ensuring that historical datasets reflect the same 
field naming conventions over a span of years can prevent 
inconsistent analyses.

\paragraph{Principle 2: Guaranteeing Accuracy and Validity in Data Entry}
Data accuracy and validity are fundamental to the utility of 
open datasets. In the NYC 311 data, issues such as 
invalid zip codes and 
nonsensical date entries (e.g., negative durations) highlight the need 
for rigorous validation processes. Governments should establish 
clear protocols for data entry, ensuring that fields are populated with 
valid, legally acceptable values. For example, fields such as 
\texttt{created\_date} and \texttt{closed\_date} should be subject to 
automated checks that prevent logical errors, such as closing an 
SR before it was opened. Ensuring the integrity of critical fields 
like geographic locations and agency names enhances the quality 
and reliability of the data. Additionally, a number of fields have 
very defined domains, such as \texttt{incident\_zip}. Such fields
should be subject to data entry validation, in this example
validating against the USPS database at the data entry point.

\paragraph{Principle 3: Optimizing Storage and Representation Efficiency}
Efficient storage and data representation are crucial for handling 
large government datasets. By eliminating redundant fields and 
encoding categorical variables, agencies can significantly reduce 
file sizes and improve the efficiency of data analysis. In the NYC 
311 dataset, for example, redundant fields such as \texttt{borough} 
and \texttt{park\_borough} could be consolidated without any data 
loss, saving space and reducing file size. Another 
example is the \texttt{taxi\_company\_borough} field, which is 99.94\% 
blank and should be evaluated to determine if its inclusion is 
necessary. Similarly, categorical variables like \texttt{complaint\_type} 
could be encoded in a standardized format to optimize storage and 
streamline analyses. Reducing file size not only saves storage space 
but also enhances the performance of queries and analysis processes.

\paragraph{Principle 4: Updating Data Dictionaries and Documentation}
Clear and accurate documentation is essential for the effective use 
of government released datasets. Our study found several 
issues where the Data Dictionary did not accurately reflect the 
actual data, such as missing field descriptions for computed fields, and 
inaccurate domain values for categorial fields such as \texttt{status}. 
As mentioned, updating the Data Dictionary to include descriptions of all fields, 
including the computed fields, e.g. \texttt{@computed\_region\_xxxx}, 
would improve transparency and usability. Regular updates should 
occur whenever there are changes in data structure, data usage, or 
content. This practice enhances the ability of external users, 
including researchers and developers, to understand and work 
effectively with the data.

\paragraph{Principle 5: Automating Quality Assurance Processes}
Automating data validation and quality assurance processes can 
significantly improve the reliability of government datasets. In the 
NYC 311 case, issues such as large spikes in SR creation and closure 
at exactly midnight (or noon) likely stem from a cron job type 
processes that distort the accurate capture of SR 
durations. By implementing real\mbox{-}time validation tools, governments 
can prevent such anomalies from entering the dataset in the first 
place. Automated 
systems can be designed to flag inconsistencies, incorrect values, 
or illogical sequences, or illogical date entries. 
Automation can also facilitate faster detection and correction of 
errors, reducing the need for manual intervention and ensuring 
the data remains clean and reliable as it is continuously updated.

\paragraph{Principle 6: Addressing Data Transparency and Accessibility}
Transparency is a core tenet of open data, and governments must ensure 
that datasets are not only accurate but also accessible and easily 
understandable. This involves providing clear metadata and making 
datasets available in user\mbox{-}friendly formats. Additionally, 
governments should communicate transparently about 
data quality issues, such as notifying the public when errors in SR 
fields (e.g., erroneous \texttt{closed\_ date(s)} are discovered, including
an explanation about how such issues will be resolved. By 
ensuring transparency and keeping the public informed, 
governments foster trust and promote wider use of datasets 
for research, civic engagement, and innovation.

\subsection{Actionable Steps}
While these principles are broad, actionable steps are required to 
implement them effectively. Governments can begin by investing in 
real\mbox{-}time validation systems, regular updates to data dictionaries, 
and tools for encoding and standardizing data fields. For example, 
systems that automate the enforcement of data validation rules 
(e.g., flagging incorrect zip codes) can significantly improve data 
quality. Collaboration with city agencies and stakeholders is crucial 
to ensure that data standards, formats, and metadata meet the needs 
of diverse users, including policymakers, researchers, and the 
general public. Regular communication between agencies and external 
users can help identify recurring issues and areas where improvements 
are needed. By facilitating regular communication and collaboration 
with various stakeholders, governments can ensure data remains relevant 
and actionable for public use.

Engagement with data scientists and statisticians is key to improving 
data curation strategies. These experts can help design and refine 
curation processes, such as creating algorithms for automatic data 
cleaning, identifying inconsistent data entries, and ensuring the 
dataset(s) structure is maintained over time. Their expertise is also 
critical in developing protocols for validating incoming data, correcting 
errors in real\mbox{-}time, and ensuring consistency across multiple data sources. 
Feedback from statisticians and data scientists, such as the issues reported 
in this paper, needs a clear and accessible channel to reach the 311 SR data 
team and other relevant city agencies. Establishing such feedback loops 
ensures that recommendations for improving data curation and validation 
are not only heard but implemented. This feedback mechanism would also allow 
ongoing dialogue between data users and city officials, leading to more dynamic 
and responsive data management practices.

A particularly valuable avenue for engagement is through events such as NYC 
Open Data Week, which fosters collaboration and innovation across sectors. 
Our involvement in this project stemmed from such an event, where we were 
inspired by the potential of open data to enhance public services. Governments 
should continue supporting these kinds of initiatives, which not only promote 
data literacy but also encourage community\mbox{-}driven improvements to datasets. 
By leveraging the expertise of data scientists, statisticians, stakeholders, 
and civic organizations, governments can ensure that their open datasets 
remain accurate, transparent, and useful for a wide range of applications.

@@

This research further emphasizes the necessity of harmonizing data
across City Agencies, particularly concerning resolving discrepancies in field
names, formats, field usage, and definitions. Consistency in data is essential for
trustworthy machine learning, where inconsistencies across sources can
lead to distorted outcomes. This challenge is even more pronounced in
datasets that span long time periods, as changes in Agency structures
or reporting standards which can introduce subtle biases. Long\mbox{-}term data
consistency is crucial for longitudinal studies and predictive
modeling, as even minor data shifts can lead to significant deviations
in model outcomes, as noted by \citet{rahm2000data} and
\citet{borgman2012conundrum}.

>>>>

Figure~\ref{fig:SRcountbyAgency} provides a breakdown of SRs by
the responsible agency, showing the cumulative 
percentage of SRs handled by each
agency over a the 2\mbox{-}year time-frame. Note that the
distribution of SRs is heavily concentrated among a few key 
agencies. The six largest agencies are:

@@

\begin{figure}[tbp]
	\centering
	\includegraphics[width = \textwidth]{SRs_by_Agency.pdf}
  	\caption{SR counts by agency with Cumulative Percentage}
	\label{fig:SRcountbyAgency}
\end{figure}

@@

We identify the presence of issues in the NYC 311 SR open dataset;
not attempting to solve them, but rather to highlight their scope, 
magnitude, and potential impact. We would recommend 
that a corrective solution should be undertaken only 
after further investigation as to the why and how such 
issues might come about, including  detailed discussions 
with the originating agency as to whether or not it is an 
actual error, correct,  or holds some other status.

@@

\paragraph{Zip Codes:}
\label{sec:zipcodesissues}
All \texttt{incident\_zip} zip codes should 
validate against the USPS database, which contains 
44,173 valid zip codes. And while the \texttt{incident\_zip} field has 
only 0.07\% invalid entries, this still results in 4,374 errors; 
not insignificant. If you group the invalid entries by 
percentage by agency, it mirrors the \textit{overall} 
breakdown of SRs by agency, potentially 
indicating a systemic problem.

@@

\paragraph{Closed before Created: Negative Duration}
Citizens, NYC Government Officials, and Agencies use the \texttt{created\_date}(s) and 
\texttt{closed\_date}(s) to measure the \textit{duration} of SRs, 
a useful surrogate to measure an agency's responsiveness. 
While duration is not directly present in the dataset, 
it can be easily computed as the difference between
\texttt{(closed\_date} and \texttt{created\_date}.  Duration is one of 
most frequently used 311metrics and is frequently analyzed, such as determining if 
one NYC borough receives faster response than another for certain 
complaints. There are 12,450 SRs where the \texttt{closed\_date} precedes the 
\texttt{created\_date}, generating nonsensical ``negative durations.'' 
Although this represents only 0.2\% of the dataset, these errors can 
significantly impact response time analyses. Eight SRs with extremely 
large negative durations ($-$730 days), all originating from the 
Department of Homeless Services (DHS); all contain an entry 
of ``1900-01-01'' as the \texttt{closed\_date}. This results 
in negative durations exceeding $-$44,601 days (or 122 years). Many
other SRs have negative durations in excess of 300+ days. Such 
anomalies, though rare, can significantly skew statistical results if 
not addressed during data cleansing. As a result, these SR rows are 
removed from the analysis.

@@

\paragraph{Identical Created Date \& Closed Date: Zero Durations}
A more prevalent issue occurs when the \texttt{closed\_date} and 
\texttt{created\_date} are exactly the same, to the second. This 
creates a \textbf{zero duration} for the SR, which is nonsensical. There 
are 163,720 zero duration SRs, representing 2.6\% of all non\mbox{-}blank data. 
99\% percent of these zero duration SRs occur within five agencies:
DHMH, DOT, DOB, DSNY, and DEP. This distribution \textbf{does not} 
mirror the overall SR distribution by agency, suggesting 
an agency\mbox{-}specific issue.

@@

\paragraph{\texttt{borough} \& \texttt{taxi\_company\_borough}:} Despite 
their names, these fields are almost entirely different, with only 
a 0.05\% match. The \texttt{taxi\_company\_borough} field is 
used exclusively by the TLC, indicating a need for consultation 
with that agency to understand the differing uses. It is 
likely that that the usage of \texttt{taxi\_company\_borough} is 
quite different than that of \texttt{borough}.

\paragraph{\texttt{agency} \& \texttt{agency\_name}:} The \texttt{agency} 
field contains the abbreviations of City agencies (e.g., NYPD, DOT, DSNY), 
while \texttt{agency\_name} contains the full names. Including both seems 
redundant, especially given the space taken up by full names in the 
CSV file. Additionally, the agency abbreviations are easily recognized.

@@

\subsection{Principles}
\paragraph{Principle 1: Ensuring Consistency Across Agencies and Time}
A key challenge in government released datasets is the 
harmonization of data fields across various Agencies and 
maintaining this consistency over time. To address this, agencies should adopt 
standardized naming conventions and formats across all datasets. 
Regular audits and data harmonization processes should be 
implemented to ensure consistency across long\mbox{-}term datasets, 
particularly as agency structures or reporting standards evolve. 
For instance, ensuring that historical datasets reflect the same 
field naming conventions over a span of years can prevent 
inconsistent analyses. These standards should be promulgated 
by the Open Data governing authority, following agreements 
by the relevant parties.

\paragraph{Principle 2: Guaranteeing Accuracy and Validity during Data Entry}
Data accuracy and validity are fundamental to the utility of 
open datasets. In the NYC 311 data, issues such as 
invalid zip codes and nonsensical date entries 
highlight the need for rigorous validation processes. Governments 
should establish clear protocols for data entry, ensuring that 
fields are populated with valid, legally acceptable values. To 
the maximum extent possible,the software should assist in 
ensuring data entries are accurate and valid. Some examples include:

\begin{itemize}[left=1.5em]
    \item Certain date fields should be promulgated by the software
    system versus manual entry. Certain dates should be populate 
    based upon an action in the system, such as a change in status.

    \item Select fields should be subjected to automated checks 
    that prevent logical errors, such as closing opening.

    \item A number of fields have clearly defined domains of legal
    values. These fields should be validated against a reference 
    dataset upon entry. 
 \end{itemize}

\paragraph{Principle 3: Optimizing Storage and Representation Efficiency}
Efficient storage and data representation are crucial for handling 
large government datasets. Our research found several duplicate and
near\mbox{-}duplicate fields in the 311 SR dataset. Eliminating 
redundant fields and encoding categorical variables can 
significantly reduce file sizes and improve the efficiency 
of data analysis. Similarly, categorical variables could be 
encoded in a standardized format to optimize storage and 
streamline analyses. Reducing file size not only saves 
storage space but also enhances the performance of queries, 
downloads,  and analysis processes.

\paragraph{Principle 4: Updated Data Dictionaries and Documentation}
Clear and accurate data documentation is essential for the effective use 
of government released datasets. Our study found several 
instances where the Data Dictionary did not accurately reflect the 
actual data. Domains of legal/allowable values should be referenced
in the Data Dictionary. Regular reviews and updates should 
occur whenever there are changes in data structure, data usage, or 
content. This practice enhances the ability of external users, 
including researchers and developers, to understand and work 
effectively with the data.

\paragraph{Principle 5: Automating Quality Assurance Processes}
Automating data validation and quality assurance processes can 
significantly improve the reliability of government datasets. Certain
unusual trends can be identified, and irregularities highlighted. 
In the NYC 311 case, issues such as large spikes in SR 
creation and closure at exactly midnight (or noon) likely 
stem from a cron job type processes that distort the 
accurate capture of SR durations. By implementing 
real\mbox{-}time and event\mbox{-}driven validation tools, governments 
can both prevent such errors from entering the dataset in the first 
place, as well as highlight anomalous trends. Automated 
systems can be designed to flag inconsistencies, incorrect values, 
illogical sequences, or illogical data entries. Automation can 
also facilitate faster detection and correction of errors, reducing 
the need for manual intervention and ensuring the data 
remains clean and reliable as it is continuously updated.

\paragraph{Principle 6: Addressing Data Transparency and Accessibility}
Transparency is a core tenet of open data, and governments must ensure 
that datasets are not only accurate but also easily accessible and 
readily understandable. This involves providing clear 
metadata and making datasets available in user\mbox{-}friendly 
formats. Additionally, governments should communicate transparently about 
data quality issues, such as notifying the public when errors in SR 
fields (e.g., erroneous \texttt{closed\_ date(s)} are discovered, including
an explanation about how such issues will be resolved. By 
ensuring transparency and keeping the public informed, 
governments foster trust and promote wider use of datasets 
for research, civic engagement, and innovation.

\subsection{Actionable Steps}
While these principles are broad, actionable steps are required to 
implement them effectively. Governments can begin by investing in 
real\mbox{-}time validation systems, regular updates to data dictionaries, 
and tools for encoding and standardizing data fields. This would include
systems that automate enforcement of data validation rules 
(e.g., flagging incorrect zip codes), significantly improving data 
quality. Collaboration with city agencies and stakeholders is crucial 
to ensure that data standards, formats, and metadata meet the needs 
of diverse users, including policymakers, researchers, and the 
general public. Regular communication between agencies and external 
users can help identify recurring issues and areas where improvements 
are needed. By facilitating regular communication and collaboration 
with various stakeholders, governments can ensure data remains relevant 
and actionable for public use.

Engagement with data scientists and statisticians is key to improving 
data curation strategies. These experts can help design and refine 
curation processes, such as creating algorithms for automatic data 
cleaning, identifying inconsistent data entries, and ensuring the 
dataset(s) structure is maintained over time. Their expertise is also 
critical in developing protocols for validating incoming data, correcting 
errors in real\mbox{-}time, and ensuring consistency across multiple data sources. 
Feedback from statisticians and data scientists, such as the issues reported 
in this paper, benefit from a clear and accessible channel to the 311 SR data 
team and other relevant city agencies. Establishing such feedback loops 
and other opportunities for public engagement, ensures that 
recommendations for improving data curation and validation 
are not only heard and discussed, but implemented. This feedback 
mechanism would also allow ongoing dialogue between 
data users and city officials, leading to more dynamic 
and responsive data management practices.

A particularly valuable avenue for engagement is through events such as NYC 
Open Data Week, which fosters collaboration and innovation across sectors. 
Our involvement in this project stemmed from such an event, where we were 
inspired by the potential of open data to enhance public services. Governments 
should continue supporting these types of initiatives, which not only promote 
data literacy, but encourage community\mbox{-}driven improvements to datasets. 
By leveraging the expertise of data scientists, statisticians, stakeholders, 
and civic organizations, governments can ensure that their open datasets 
remain accurate, transparent, and useful for a wide range of applications.

@@

This research further emphasizes the necessity of harmonizing data
across City agencies, particularly resolving discrepancies in field
names, formats, usage, and definitions. Consistency in data is essential for
trustworthy machine learning, where inconsistencies across sources can
lead to distorted outcomes. This challenge is even more pronounced in
datasets that span long time periods, as changes in agency structures
or reporting standards which can introduce subtle biases. Long\mbox{-}term data
consistency is crucial for longitudinal studies and predictive
modeling, as even minor data shifts can lead to significant deviations
in model outcomes, as noted by \citet{rahm2000data} and
\citet{borgman2012conundrum}.
